---
title: "Aggregate Analysis"
output:
  html_document:
    df_print: paged
---

## Load Packages

```{r, message = F, warning = F}
library(tidyverse)
library(here)
library(janitor)
library(dataedu)
```

## Import Data

```{r}
race_pdf <-
  dataedu::race_pdf
```


We create a data frame out of the imported pdf.

```{r, eval = FALSE}
race_df <-
  race_pdf %>%
  # Turn each page into a tibble
  map(~ as_tibble(.x, .name_repair = "unique")) %>% 
  # Make data frame and remove unnecessary rows
  map_df(~ slice(.,-1:-2)) %>%
  # Use descriptive column names
  set_names(
    c(
      "school_group",
      "school_name",
      "grade",
      "na_num", # Native American number of students
      "na_pct", # Native American percentage of students
      "aa_num", # African American number of students
      "aa_pct", # African American percentage
      "as_num", # Asian number of students
      "as_pct", # Asian percentage
      "hi_num", # Hispanic number of students
      "hi_pct", # Hispanic percentage
      "wh_num", # White number of students
      "wh_pct", # White percentage
      "pi_pct", # Pacific Islander percentage
      "blank_col",
      "tot" # Total number of students (from the Race PDF)
    )
  )

view(race_df)
```

Now we want to clean the dataset.

```{r, eval = FALSE}
race_df2 <-
  race_df %>%
  # Remove unnecessary columns
  select(-school_group, -grade, -pi_pct, -blank_col) %>%
  # Filter to get grade-level numbers
  filter(str_detect(school_name, "Total"),
         school_name != "Grand Total") %>%
  # Clean up school names
  mutate(school_name = str_replace(school_name, "Total", "")) %>%
  # Remove white space
  mutate_if(is.character, trimws) %>%
  # Turn percentage columns into numeric and decimal format
  mutate_at(vars(contains("pct")), list( ~ as.numeric(str_replace(., "%", "")) / 100))

view(race_df2)
```

We import the Free Reduced Price Lunch pdf. 

```{r}
frpl_pdf <-
  dataedu::frpl_pdf
```

Similarly to what we've done with the race pdf, we change the names of the columns to something more understandable.


```{r, eval=FALSE}
frpl_df <-
  frpl_pdf %>%
  # Turn each page into a tibble
  map(~ as_tibble(.x, .name_repair = "unique")) %>% 
  # Make data frame and remove unnecessary rows
  map_df( ~ slice(.,-1)) %>%
  # Use descriptive column names
  set_names(
    c(
      "school_name",
      "not_eligible_num", # Number of non-eligible students,
      "reduce_num", # Number of students receiving reduced price lunch
      "free_num",   # Number of students receiving free lunch
      "frpl_num",  # Total number of students (from the FRPL PDF)
      "frpl_pct" # Free/reduced price lunch percentage
    )
  )

```

We remove blank rows and the aggregations that are not in district level. 

```{r, eval = FALSE}
frpl_df2 <-
  frpl_df %>%
  filter(
    # Remove blanks
    school_name != "",
    # Filter out the rows in this list
    !school_name %in% c(
      "ELM K_08",
      "Mid Schl",
      "High Schl",
      "Alt HS",
      "Spec Ed Total",
      "Cont Alt Total",
      "Hospital Sites Total",
      "Dist Total"
    )
  ) %>%
  # Turn percentage columns into numeric and decimal format
  mutate(frpl_pct = as.numeric(str_replace(frpl_pct, "%", "")) / 100)
```

Join two data frames that we just created by school name. 

```{r, eval = FALSE}
# create full dataset, joined by school name
joined_df <-
  left_join(race_df2, frpl_df2, by = c("school_name")) %>%
  mutate_at(2:17, as.numeric)

view(joined_df)
```


## RQ:What is the racial distribution in low poverty schools?

1.  We want to calculate, for each race, the number of students in 'low
    poverty' schools. This is defined by NCES as schools that are over 25% FRPL
    [@ncesfrpl]. When a school is over 75% FRPL, we count the number of students
    for that particular race under the variable `[racename]_povnum`.
2.  The {janitor} package has a handy `adorn_totals()` function that sums
    columns for you. This is important because we want a weighted average of
    students in each category, so we need the total number of students in each
    group.
3.  We create the weighted average of the percentage of each race by dividing
    the number of students by race by the total number of students.
4.  To get FRPL percentage for all schools, we have to recalculate `frpl_pct` (otherwise, it would not be a weighted average).
5.  To calculate the percentage of students by race who are in high poverty
    schools, we must divide the number of students in high poverty schools by
    the total number of students in that race.
    
```{r}
district_merged_df2 <-
  joined_df %>%
  # Calculate low poverty numbers
  mutate(
    hi_povnum = case_when(frpl_pct < .25 ~ hi_num),
    aa_povnum = case_when(frpl_pct < .25 ~ aa_num),
    wh_povnum = case_when(frpl_pct < .25 ~ wh_num),
    as_povnum = case_when(frpl_pct < .25 ~ as_num),
    na_povnum = case_when(frpl_pct < .25 ~ na_num)
  ) %>%
  adorn_totals() %>%
  # Create percentage by demographic
  mutate(
    na_pct = na_num / tot,
    aa_pct = aa_num / tot,
    as_pct = as_num / tot,
    hi_pct = hi_num / tot,
    wh_pct = wh_num / tot,
    frpl_pct = (free_num + reduce_num) / frpl_num,
    # Create percentage by demographic and poverty
    hi_povsch = hi_povnum / hi_num[which(school_name == "Total")],
    aa_povsch = aa_povnum / aa_num[which(school_name == "Total")],
    as_povsch = as_povnum / as_num[which(school_name == "Total")],
    wh_povsch = wh_povnum / wh_num[which(school_name == "Total")],
    na_povsch = na_povnum / na_num[which(school_name == "Total")]
  )
```

Put the data in a tidy format for plotting.
```{r}
district_tidy_df2 <-
  district_merged_df2 %>%
  pivot_longer(
    cols = -matches("school_name"),
    names_to = "category",
    values_to = "value"
  )
```

Now we can plot the racial distribution of low poverty schools. 
```{r}
district_tidy_df2 %>%
  filter(school_name == "Total",
         str_detect(category, "povsch")) %>%
  ggplot(aes(x = reorder(category,-value), y = value)) +
  geom_bar(stat = "identity", aes(fill = factor(category))) +
  geom_text(aes(label = scales::percent(value)), vjust = -0.5, size = 3) +  # Add text labels
  labs(title = "Distribution of Subgroups in Low Poverty Schools",
       x = "Subgroup",
       y = "Percentage in Low Poverty Schools") +
  scale_x_discrete(
    labels = c(
      "aa_povsch" = "Black",
      "wh_povsch" = "White",
      "hi_povsch" = "Hispanic",
      "as_povsch" = "Asian",
      "na_povsch" = "Native Am."
    )
  ) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_dataedu() +
  theme_dataedu() +
  theme(legend.position = "none")
```