---
title: "Gradebook Analysis"
author: "Ha Thuong Tran"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r echo = TRUE, message = FALSE, warning = FALSE}
# Load libraries
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(dataedu)
library(corrplot)
library(car)
library("olsrr")
```

## Import data
```{r eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE}
# See the current working directory
getwd()
```

```{r read in data, eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}
# Use readxl package to read and import file and assign it a name
ExcelGradeBook <-
  read_excel(
    here("ExcelGradeBook.xlsx"),
    sheet = 1,
    skip = 10
  )
```

```{r assign to new name, echo = TRUE, message = FALSE, warning = FALSE}
# Rename data frame
gradebook <- ExcelGradeBook
```


## Process data 

```{r echo = TRUE, message = FALSE, warning = FALSE}
#Rename the columns 
gradebook <- 
  gradebook %>% 
  clean_names()
```

```{r}
# Remove rows that contain nothing but NA values 
gradebook <- gradebook[rowSums(is.na(gradebook)) < ncol(gradebook), ]

# Remove unneeded columns  
unneeded_columns <- c("repeated_grades", "make_your_own_categories", "another_type_2", "assessment_insert_new_columns_before_here")
gradebook <- gradebook[, !(names(gradebook) %in% unneeded_columns)]

```


## Data analysis

### RQ1: What is the relationship between formative and summative assessments?

**Linearity**
```{r message = FALSE, warning = FALSE, results = 'hide', fig.showtext = TRUE}  
# Scatter plot between formative assessment and summative assessments
# To determine linear relationship
# With line of best fit
gradebook %>%
  ggplot(aes(x = formative_assessments,
             y = summative_assessments)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm",
              se = TRUE) +
  labs(title = "Relationship Between Summative Assessments and Formative Assessments (with Line of Best Fit)",
       x = "Formative Assessment Score",
       y = "Summative Assessment Score") +
  theme_dataedu()
```
**Normal distribution**

```{r}
column_to_check <- gradebook$summative_assessments

# Shapiro-Wilk normality test
shapiro.test(column_to_check)

```
Since the p-value is greater than 0.05, the null hypothesis is not rejected, suggesting that the data may be normally distributed.

**Outliers**

```{r message = FALSE, warning = FALSE, results = 'hide', fig.cap = "Distribution of Formative Assessment Scores", fig.showtext = FALSE}
# Boxplot of formative assessment scores
# To determine if there are any outliers
gradebook %>%
  ggplot(aes(x = "",
             y = formative_assessments)) +
  geom_boxplot(fill = dataedu_colors("yellow")) +
  labs(title = "Distribution of Formative Assessment Scores",
       x = "Formative Assessment",
       y = "Score") +
  theme_dataedu()

# Assuming 'gradebook' is your dataset
# Load necessary libraries if not already loaded
# install.packages(c("ggplot2", "tidyr"))
library(ggplot2)
library(tidyr)

# Melt the dataset using tidyr::pivot_longer
melted_gradebook <- pivot_longer(gradebook, 
                                 cols = c("homeworks", "classworks", "projects", "summative_assessments", "formative_assessments", "running_average"), 
                                 names_to = "Assessment_Type", 
                                 values_to = "Score")

melted_gradebook$Assessment_Type <- factor(melted_gradebook$Assessment_Type,        levels = c("homeworks", "classworks", "projects", "summative_assessments", "formative_assessments", "running_average"))                                      
# Create box plots for each variable
ggplot(melted_gradebook, aes(x = Assessment_Type, y = Score)) +
  geom_boxplot(fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Box Plots of Various Assessments in Gradebook", x = "Assessment Type", y = "Score") +
  theme_minimal() +
  scale_x_discrete(labels = c(
    "homeworks" = "Homeworks",
    "classworks" = "Classworks",
    "projects" = "Projects",
    "summative_assessments" = "Summative",
    "formative_assessments" = "Formative",
    "running_average" = "Average"
))
```

```{r message = FALSE, warning = FALSE, results = 'hide', fig.cap = "Distribution of Summative Assessment Scores", fig.showtext = FALSE}
# Boxplot of summative assessment scores
# To determine if there are any outliers
gradebook %>%
  ggplot(aes(x = "",
             y = summative_assessments)) +
  geom_boxplot(fill = dataedu_colors("yellow")) +
  labs(title = "Distribution of Summative Assessment Scores",
       x = "Summative Assessment",
       y = "Score") +
  theme_dataedu()
```

We remove the outliers in summative assessment because Pearson Correlation Coefficient is sensitive to outliers.

```{r}
# Extract the column
summative_assessments <- gradebook$summative_assessments

# Calculate the first and third quartiles
Q1 <- quantile(summative_assessments, 0.25)
Q3 <- quantile(summative_assessments, 0.75)

# Calculate the interquartile range (IQR)
IQR_value <- IQR(summative_assessments)

# Define the lower and upper bounds for outliers
lower_bound <- Q1 - 1.5 * IQR_value
upper_bound <- Q3 + 1.5 * IQR_value

# Identify outliers
outliers <- gradebook$summative_assessments[gradebook$summative_assessments < lower_bound | gradebook$summative_assessments > upper_bound]

# Remove outliers
gradebook_no_outliers_summative <- gradebook[gradebook$summative_assessments >= lower_bound & gradebook$summative_assessments <= upper_bound, ]

# To determine if there are any outliers
boxplot(gradebook_no_outliers_summative$summative_assessments, main="Boxplot of Summative Assessments After Outlier Removal")

view(outliers)
view(gradebook_no_outliers_summative)

```

**Correlation Analysis**
```{r}
# Compute Pearson correlation
correlation_coefficient <- cor(gradebook_no_outliers_summative$formative_assessments, gradebook_no_outliers_summative$summative_assessments)

# Assuming x and y are your variables
cor_test_result <- cor.test(gradebook_no_outliers_summative$formative_assessments, gradebook_no_outliers_summative$summative_assessments)

# Access correlation coefficient
cor_coefficient <- cor_test_result$estimate

# Access p-value
p_value <- cor_test_result$p.value

# Print the results
print(paste("Correlation Coefficient (r):", cor_coefficient))
print(paste("P-value:", p_value))


# Print the result
cat("Pearson correlation coefficient between formative and summative assessments:", correlation_coefficient, "\n")

```

### RQ2: What is the relationship between homework, classwork, and project grades? 

**Linearity**
Already did in RQ3

**Normal distribution**
```{r}
# Shapiro-Wilk normality test
shapiro.test(gradebook$classworks)
shapiro.test(gradebook$homeworks)
shapiro.test(gradebook$projects)

```

**Remove outliers**
Already did in RQ3.

**Calculate correlation matrix**
```{r}
my_data <- gradebook[, c('classworks', 'homeworks', 'projects')]
library("Hmisc")
res2 <- rcorr(as.matrix(my_data))
res2


```

### RQ3: What has a stronger impact on the final grade: homework, classwork, or project grades?

**Linearity**
```{r message = FALSE, warning = FALSE, results = 'hide', fig.showtext = TRUE}  

# Scatter plot between homeworks and grades by percent
gradebook %>%
  ggplot(aes(x = homeworks,
             y = running_average)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm",
              se = TRUE) +
  labs(title = "Relationship Between Homeworks and Final Grade (with Line of Best Fit)",
       x = "Homework Score",
       y = "Final Grade") +
  theme_dataedu()

# Scatter plot between classworks and grades by percent
gradebook %>%
  ggplot(aes(x = classworks,
             y = running_average)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm",
              se = TRUE) +
  labs(title = "Relationship Between Classworks and Final Grade (with Line of Best Fit)",
       x = "Classwork Score",
       y = "Final Grade") +
  theme_dataedu()

# Scatter plot between projects and grades by percent
gradebook %>%
  ggplot(aes(x = projects,
             y = running_average)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm",
              se = TRUE) +
  labs(title = "Relationship Between Projects and Final Grade (with Line of Best Fit)",
       x = "Project Score",
       y = "Final Grade") +
  theme_dataedu()
```

**Homoscedasticity**

```{r}
# Perform Breusch-Pagan for classwork 
model_classwork <- lm(classworks ~ running_average, data = gradebook)
bp_test_classwork <- ncvTest(model_classwork)
print(bp_test_classwork)

# Perform Breusch-Pagan for homework 
model_homework <- lm(homeworks ~ running_average, data = gradebook)
bp_test_homework <- ncvTest(model_homework)
print(bp_test_homework)

# Perform Breusch-Pagan for project 
model_project <- lm(projects ~ running_average, data = gradebook)
bp_test_project <- ncvTest(model_project)
print(bp_test_project)
```
All of the p-values are greater than 0.05, so we can assume homoscedasticity.

**Multicollinearity**

```{r}
#define multiple linear regression model
model <- lm(running_average ~ classworks + homeworks + projects, data=gradebook)

#calculate the VIF for each predictor variable in the model
vif(model)
```
Since each of the VIF values for the predictor variables in the model are close to 1, multicollinearity is not a problem in the model.

**Outliers**
```{r}
# check for outliers in classwork
gradebook %>%
  ggplot(aes(x = "",
             y = classworks)) +
  geom_boxplot(fill = dataedu_colors("yellow")) +
  labs(title = "Distribution of Classwork Scores",
       x = "Classwork",
       y = "Score") +
  theme_dataedu()

# check for outliers in homework
gradebook %>%
  ggplot(aes(x = "",
             y = homeworks)) +
  geom_boxplot(fill = dataedu_colors("yellow")) +
  labs(title = "Distribution of Homework Scores",
       x = "Homework",
       y = "Score") +
  theme_dataedu()

# check for outliers in projects
gradebook %>%
  ggplot(aes(x = "",
             y = projects)) +
  geom_boxplot(fill = dataedu_colors("yellow")) +
  labs(title = "Distribution of Project Scores",
       x = "Project",
       y = "Score") +
  theme_dataedu()
```
No outliers are present. 

**Normal distributions of residuals**
```{r}
# Perform normality test to check for normal distributions of residuals 
ols_test_normality(model_classwork)
ols_test_normality(model_homework)
ols_test_normality(model_project)
```
Since all of the p-values are greater than 0.05, we fail to reject the null hypothesis and assume normal distribution of residuals.

**Compute the model coefficients**
```{r}
summary(model)
```
All of the independent variables are significant positive predictors of final grade. Classwork scores are the most important, followed by homework scores and project scores.

### RQ4: Better late or absent? What is the impact of being late vs. being absent?

**Simulate data**

First, we need to simulate some data. Since we already have some data in these columns, we fit a zero-inflated negative binomial model to the existing data, then simulate new data based on the count distribution of the existing data.  

```{r}
# Replace "no match" with NA in the 'late' column
gradebook$late[gradebook$late == "No match"] <- NA

# Replace "no match" with NA in the 'absent' column
gradebook$absent[gradebook$absent == "No match"] <- NA

library(pscl)

# Identify the rows with original data
original_data_late <- !is.na(gradebook$late)
original_data_absent <- !is.na(gradebook$absent)

# Fit a zero-inflated negative binomial model to the existing data
fit_late <- zeroinfl(late ~ 1, dist = "negbin", data = gradebook)
fit_absent <- zeroinfl(absent ~ 1, dist = "negbin", data = gradebook)

# Get the probability of zero-inflation
p_late <- predict(fit_late, type = "zero")
p_absent <- predict(fit_absent, type = "zero")

# Get the mean of the count distribution
lambda_late <- predict(fit_late, type = "count")
lambda_absent <- predict(fit_absent, type = "count")

# Simulate the distribution for new data in the remaining rows
gradebook$simulated_late <- gradebook$late  # Create a new column with the original data
gradebook$simulated_absent <- gradebook$absent  # Create a new column with the original data

# Simulate late and absent data for rows where there's no original data
set.seed(1)

gradebook$simulated_late[!original_data_late] <- ifelse(rbinom(sum(!original_data_late), size = 1, prob = p_late) > 0, 0, rpois(sum(!original_data_late), lambda = lambda_late))

gradebook$simulated_absent[!original_data_absent] <- ifelse(rbinom(sum(!original_data_absent), size = 1, prob = p_absent) > 0, 0, rpois(sum(!original_data_absent), lambda = lambda_absent))

# Convert simulated late and absent data to numeric and store them in columnes "late" and "absent" 
gradebook$late <- as.numeric(gradebook$simulated_late)
gradebook$absent <- as.numeric(gradebook$simulated_absent)
```

**Create bar charts for 'late' and 'absent'**

```{r}
library(ggplot2)

# Bar chart for 'late'
ggplot(gradebook, aes(x = late)) +
  geom_bar(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Late Counts", x = "Late Count", y = "Frequency") +
  theme_minimal()

# Bar chart for 'absent'
ggplot(gradebook, aes(x = absent)) +
  geom_bar(fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Absent Counts", x = "Absent Count", y = "Frequency") +
  theme_minimal()
```


**Homoscedasticity**
```{r}

model_late_absent <- lm(running_average ~ late + absent, data=gradebook)
residuals <- residuals(model_late_absent)
plot(model_late_absent, which = 1)

```
**Multicollinearity**
```{r}
car::vif(model_late_absent)
```

**Normal distribution of errors**
```{r}
ols_test_normality(model_late_absent)
```

**Run MLR model**
```{r}
summary(model_late_absent)
```

### RQ5: What has a stronger impact on final grade: race or gender?

**Generate random data**
```{r}
set.seed(1)

# Generate random gender data
gradebook$gender <- sample(c("Male", "Female"), nrow(gradebook), replace = TRUE)

#Generate random race data 
gradebook$race <- sample(c("Caucasian", "Asian", "Black"), nrow(gradebook), replace = TRUE)

# Print the first few rows of the data frame
head(gradebook)

```

**Visualize data**

```{r}
library(ggplot2)

# Pie chart for gender
ggplot(gradebook, aes(x = "", fill = gender)) +
  geom_bar(width = 1, stat = "count") +
  geom_text(aes(label = stat(count), group = gender), stat = "count", position = position_stack(vjust = 0.5), color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Gender Distribution")

# Pie chart for race
ggplot(gradebook, aes(x = "", fill = race)) +
  geom_bar(width = 1, stat = "count") +
  geom_text(aes(label = stat(count), group = race), stat = "count", position = position_stack(vjust = 0.5), color = "white") +
  coord_polar(theta = "y") +
  labs(title = "Race Distribution")



```

**Homoscedasticity**
```{r}

model_race_gender <- lm(running_average ~ race + gender, data=gradebook)
residuals <- residuals(model_race_gender)
plot(model_race_gender, which = 1)

```
**Multicollinearity**
```{r}
car::vif(model_race_gender)
```
**Normal distribution of errors**
```{r}
ols_test_normality(model_race_gender)
```
**Run MLR**
```{r}
summary(model_race_gender)
```
```{r}
model_race <- lm(running_average ~ race, data = gradebook)
summary(model_race)

model_gender <- lm(running_average ~ gender, data = gradebook)
summary(model_gender)
```

### RQ6: What has a stronger impact on final grade: age or financial status?

**Generate random data**
```{r}
set.seed(1)

# Simulate the 'age' column using a normal distribution
gradebook$age <- round(rnorm(nrow(gradebook), mean = 25, sd = 5))

# Simulate the 'financial_status' column using a normal distribution for scores
gradebook$financial_status <- round(rnorm(nrow(gradebook), mean = 75, sd = 10))
```

**Plot histograms for age and financial status**

```{r}
library(ggplot2)

# Histogram for 'age'
ggplot(gradebook, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age", x = "Age", y = "Frequency") +
  theme_minimal()

# Histogram for 'financial_status'
ggplot(gradebook, aes(x = financial_status)) +
  geom_histogram(binwidth = 5, fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Financial Status", x = "Financial Status", y = "Frequency") +
  theme_minimal()
```

**Linearity**
```{r message = FALSE, warning = FALSE, results = 'hide', fig.showtext = TRUE}  

# Scatter plot between age and grades by percent
gradebook %>%
  ggplot(aes(x = age,
             y = running_average)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Relationship Between Age and Final Grade (with Line of Best Fit)",
       x = "Age",
       y = "Final Grade") +
  theme_dataedu()

# Scatter plot between financial status and grades by percent
gradebook %>%
  ggplot(aes(x = financial_status,
             y = running_average)) +
  geom_point(color = dataedu_colors("green")) +
  geom_smooth(method = "lm",
              se = TRUE) +
  labs(title = "Relationship Between Financial Status and Final Grade (with Line of Best Fit)",
       x = "Financial Status",
       y = "Final Grade") +
  theme_dataedu()
```
**Loess curve**

```{r}
plot(gradebook$age,gradebook$running_average)
lines(loess.smooth(gradebook$age, gradebook$running_average), col = "red")


plot(gradebook$financial_status,gradebook$running_average)
lines(loess.smooth(gradebook$financial_status, gradebook$running_average), col = "red")
```

```{r}
model <- lm(running_average ~ age + financial_status, data = gradebook)
```

**Homoscedasticity**
```{r}

model_age_finance <- lm(running_average ~ age + financial_status, data=gradebook)
residuals <- residuals(model_age_finance)
plot(model_age_finance, which = 1)

```
**Multicollinearity**
```{r}
vif(model_age_finance)
```

**Normal distribution of errors**
```{r}
ols_test_normality(model_age_finance)
```

**Run MLR**
```{r}
summary(model_age_finance)
```

